---
title: "Assignment 10: Data Scraping"
author: "Esther Yang"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Rename this file `<FirstLast>_A10_DataScraping.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up 
1. Set up your session:

* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Check your working directory

```{r 1. Project Setup, message = FALSE}
#1 
library(tidyverse)
library(rvest)
getwd()

```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2024 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2024>
 
Indicate this website as the as the URL to be scraped. (In other words, read the contents into an `rvest` webpage object.)

```{r set.the.scraping.website}
#2 
url <- "https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2024"
webpage <- read_html(url)
```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * PWSID
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Maximum Day Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to four separate variables.

>HINT: The first value should be "Durham", the second "03-32-010", the third "Municipality", and the last should be a vector of 12 numeric values (represented as strings)".

```{r scrape.the.data}

#3 
water_system_name <- webpage %>%
  html_element(xpath = "//td[contains(.,'Water System Name')]/following-sibling::td") %>%
  html_text(trim = TRUE)

pwsid <- webpage %>%
  html_element(xpath = "//td[contains(.,'PWSID')]/following-sibling::td") %>%
  html_text(trim = TRUE)

ownership <- webpage %>%
  html_element(xpath = "//td[contains(.,'Ownership')]/following-sibling::td") %>%
  html_text(trim = TRUE)

tables <- webpage %>% html_elements("table")
max_table <- tables[[13]] %>% html_table(fill = TRUE)

max_cols_idx <- which(names(max_table) == "Max DayUse (MGD)")   
if(length(max_cols_idx) == 0){
  stop("No columns named 'Max DayUse (MGD)' found in max_table. Run names(max_table) to inspect.")
}
max_vals_raw <- unlist(max_table[, max_cols_idx], use.names = FALSE)
max_day_use <- max_vals_raw %>%
  as.character() %>%        
  stringr::str_trim() %>%   
  { gsub("[^0-9\\.-]", "", .) } %>%
  as.numeric() %>%
  na.omit() %>%
  as.character()

water_system_name
pwsid
ownership
max_day_use
```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in chronological order. You can overcome this by creating a month column manually assigning values in the order the data are scraped: "Jan", "May", "Sept", "Feb", etc... 

5. Create a line plot of the maximum daily withdrawals across the months for 2024, making sure, the months are presented in proper sequence.

```{r create.a.dataframe.from.scraped.data}
#4 
months <- c("Jan","Feb","Mar","Apr","May","Jun","Jul","Aug","Sep","Oct","Nov","Dec")

water_data <- tibble(
  Month = factor(months, levels = months),
  Max_Day_Use_MGD = as.numeric(max_day_use),
  Water_System = rep(water_system_name, 12),
  PWSID = rep(pwsid, 12),
  Ownership = rep(ownership, 12),
  Year = rep(2024, 12),
  Date = as.Date(paste0("2024-", months, "-01"), format = "%Y-%b-%d")
)

#5 
ggplot(water_data, aes(x = Month, y = Max_Day_Use_MGD, group = 1)) +
  geom_line(color = "blue", size = 1) +
  geom_point(color = "red", size = 2) +
  labs(title = "Durham 2024 Max Daily Withdrawals",
       x = "Month", y = "Max Day Use (MGD)") +
  theme_minimal()

```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function with two inputs - "PWSID" and "year" - that:
  - Creates a URL pointing to the LWSP for that PWSID for the given year
  - Creates a website object and scrapes the data from that object (just as you did above)
  - Constructs a dataframe from the scraped data, mostly as you did above, but includes the PWSID and year provided as function inputs in the dataframe. 
  - Returns the dataframe as the function's output


```{r construct.a.scraping.function}
#6. 
get_LWSP_data <- function(PWSID, year) {
  url <- paste0("https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=", PWSID, "&year=", year)
  page <- read_html(url)
  
  system_name <- page %>%
    html_element(xpath = "//td[contains(.,'Water System Name')]/following-sibling::td") %>%
    html_text(trim = TRUE)
  
  ownership <- page %>%
    html_element(xpath = "//td[contains(.,'Ownership')]/following-sibling::td") %>%
    html_text(trim = TRUE)
  
  tables <- page %>% html_elements("table") %>% map(html_table, fill = TRUE)
  
  tbl_idx <- which(map_lgl(tables, ~ any(grepl("(?i)max.*day.*(use|withdr)", names(.), perl = TRUE))))
  if (length(tbl_idx) == 0) stop("No table with 'Max Day Use/Withdrawal' found for this system/year.")
  
  monthly_tbl <- tables[[tbl_idx[1]]]
  idx <- which(grepl("(?i)max.*day.*(use|withdr)", names(monthly_tbl), perl = TRUE))
  
  vals <- unlist(monthly_tbl[, idx], use.names = FALSE)
  
  max_day_use <- vals %>%
    as.character() %>%
    stringr::str_trim() %>%
    gsub(",", "", .) %>%
    gsub("[^0-9\\.]", "", .) %>%
    as.numeric() %>%
    na.omit()
  
  if (length(max_day_use) > 12) max_day_use <- head(max_day_use, 12)
  if (length(max_day_use) < 12) max_day_use <- c(max_day_use, rep(NA, 12 - length(max_day_use)))
  
  months <- c("Jan","Feb","Mar","Apr","May","Jun","Jul","Aug","Sep","Oct","Nov","Dec")
  
  df <- tibble(
    Month = factor(months, levels = months),
    Max_Day_Use_MGD = as.numeric(max_day_use),
    Water_System = system_name,
    PWSID = PWSID,
    Ownership = ownership,
    Year = year,
    Date = as.Date(paste0(year, "-", months, "-01"), format = "%Y-%b-%d")
  )
  
  return(df)
}
```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2020. Then show the structure of the dataframe with either the `str()` or the `glimpse()` function. 

```{r fetch.and.plot.Durham.2020.data}
#7 

durham2020 <- get_LWSP_data("03-32-010", 2020)
ggplot(durham2020, aes(x = Month, y = Max_Day_Use_MGD, group = 1)) +
  geom_line(color = "blue", size = 1) +
  geom_point(color = "red", size = 2) +
  labs(title = "Durham 2020 Max Daily Withdrawals",
       x = "Month", y = "Max Day Use (MGD)") +
  theme_minimal()

glimpse(durham2020)

```

8. Use the function above to extract data for Cary (PWSID = '03-32-010') in 2020. Combine this data with the Durham data collected above and create a plot that compares Cary's to Durham's water withdrawals.

```{r fetch.and.plot.Cary.2020.data}
#8 
cary2020 <- get_LWSP_data("03-92-020", 2020)
compare2020 <- bind_rows(durham2020, cary2020)

ggplot(compare2020, aes(x = Month, y = Max_Day_Use_MGD, color = Water_System, group = Water_System)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  labs(title = "2020 Max Daily Withdrawals Cary vs Durham",
       x = "Month", y = "Max Day Use (MGD)",
       color = "System") +
  theme_minimal()


```


9. Use the code & function you created above to plot Cary's max daily withdrawal by months for the years 2018 thru 2023.Add a smoothed line to the plot (method = 'loess').

> TIP: See Section 3.2 in the "10_Data_Scraping.Rmd" where we apply "map2()" to iteratively run a function over two inputs. Pipe the output of the map2() function to `bind_rows()` to combine the dataframes into a single one, and use that to construct your plot.

```{r fetch.and.plot.Cary.multiyear.data }
#9 
years <- 2018:2023
pwsid_cary <- "03-92-020"

map2(rep(pwsid_cary, length(years)), years, get_LWSP_data) %>%
  bind_rows() %>%
  ggplot(aes(x = Month, y = Max_Day_Use_MGD, color = factor(Year), group = Year)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  geom_smooth(aes(group = 1), method = "loess", se = FALSE,
              color = "black", linetype = "dashed", size = 1) +
  labs(title = "Cary Max Daily Withdrawals for the years 2018 thru 2023",
       x = "Month",
       y = "Max Day Use (MGD)",
       color = "Year") +
  theme_minimal()
```

Question: Just by looking at the plot (i.e. not running statistics), does Cary have a trend in water usage over time?
> Answer:Cary's maximum daily withdrawals appear relatively consistent across years from 2018 to 2023. While there is some seasonal variation of peaking around June to July and dropping in winter, the yearly lines largely overlap, suggesting no strong upward or downward trend in overall water usage over time.
>
